---
output: html_document
---

# Exploratory Data Analysis in R

Resource: https://r4ds.had.co.nz/tidy-data.html 

```{r setup, include=FALSE}
# Load libraries
library(ggplot2)
library(here)
library(skimr)
library(tidyverse)

# Set global theme for plots
theme_set(theme_minimal())
```

## Import data

`here` makes it easier to work with file paths by automatically detecting your project's root directory, so you don't have to worry about working directory issues or writing complex relative paths.

Resource: <https://here.r-lib.org>

```{r import data}
# import the file using here and read_csv, expecting 89184 obs. of 84 variables
survey_data = readr::read_csv(here::here("data", "survey_results_public.csv")) %>% janitor::clean_names()
```

## View data

The following are some ways to examine the data!

```{r glimpse}
# quick peek
glimpse(survey_data)
```

```{r skim}
# with a little more detail
skimr::skim(survey_data)
```

```{r head}
# view the first few rows, default is 6
head(survey_data)
```

```{r tail}
# view the last few rows, default is 6
tail(survey_data)
```

```{r colnames}
# view column names
colnames(survey_data)
```

```{r number of rows}
# count number of rows
nrow(survey_data)
```

### Exercise 1

1.  Read your data into R and view it

## Select data

`dplyr` is part of the larger `tidyverse`, it provides convenient tools for the most common data manipulation tasks e.g., selecting rows/columns and filtering.

Resource: <https://cengel.github.io/R-data-wrangling/dplyr.html>

### Select column or row by position

```{r single row}
# select single row based on position
survey_data %>% 
  dplyr::slice(2) # R uses 1-based indexing
```

```{r single column}
# select single column based on position
survey_data %>% 
  dplyr::select(2)
```

```{r range of rows}
# select a range of rows based on position
survey_data %>% 
  dplyr::slice(2:5)
```

```{r specific rows}
# select specific rows based on position
survey_data %>% 
  dplyr::slice(c(2, 6, 11))
```

```{r range of columns}
# select a range of columns based on position
survey_data %>% 
  dplyr::select(2:5)
```

```{r specific cols}
# select specific columns based on position
survey_data %>% 
  dplyr::select(c(2, 6, 11))
```

```{r negative index}
# select last 10 rows
survey_data %>% 
  dplyr::slice_tail(n = 10)
```

### Select column by name

```{r}
# Select specific column by name 
survey_data %>% 
  dplyr:: select(RemoteWork)
```

## Filtering Data

Different ways to identify and count specific values/categories

### Specific conditions

```{r filter numeric}
# Only grab those with comp total greater than 500, 000
survey_data %>%
  filter(comp_total > 500000)
```

### Specific cateogries

```{r filter category}
# Only grab those from Australia
survey_data %>%
  filter(country == "Australia")
```

### Multiple conditions

```{r filter multiple}
# Only grab those from Australia with a comp total greater than 500, 0000
survey_data %>%
  filter(comp_total > 500000 & country == "Australia")
```

## Plotting Data

`ggplot2` is part of the `tidyverse` and it is a package that is dedicated to data visualization.

Resources:

-   <https://ggplot2.tidyverse.org/articles/ggplot2.html>
-   <https://r-graph-gallery.com/ggplot2-package.html>
-   <https://exts.ggplot2.tidyverse.org/gallery/>

```{r plot WorkExp}
# Start with basic histogram
ggplot(survey_data, aes(x = work_exp)) +
  geom_histogram(bins = 30, na.rm = TRUE)

# # Add colors and labels
# ggplot(survey_data, aes(x = WorkExp)) +
#   geom_histogram(bins = 30, 
#                 na.rm = TRUE,
#                 fill = "steelblue",
#                 color = "white") +
#   labs(title = "Distribution of Work Experience",
#        subtitle = "Stack Overflow Developer Survey 2023",
#        x = "Years of Work Experience",
#        y = "Count")
```

### Exercise 2

1.  Plot your own data
2.  Choose single column to plot

# Lunch Break 

## Reshaping data

Common problems: 
- Variable across multiple columns 
- Observations across multiple rows

Resource: https://cran.r-project.org/web/packages/tidyr/vignettes/pivot.html 

### Pivot longer

https://tidyr.tidyverse.org/reference/pivot_longer.html 

```{r pivot longer}
# making data longer by converting columns to rows
survey_data_long <- survey_data %>%
  pivot_longer(cols = -response_id, # keep ResponseId as identifier
               names_to = "variable",
               values_to = "value",
               values_transform = as.character) #need to correct format

# pivot_longer(survey_data, 
#              c('CompTotal', 'ConvertedCompYearly'), 
#              names_to = 'money', 
#              values_to = 'dollah')

dim(survey_data_long)
```

### Pivot wider

https://tidyr.tidyverse.org/reference/pivot_wider.html

```{r pivot wider}
survey_data_wide <- survey_data %>%
  pivot_wider(id_cols = response_id,
              names_from = age,
              values_from = -response_id)

dim(survey_data_wide)
```

### Exercise 3

1.  Convert your own data into a tidier format.

## Clean data

### Missing Data

```{r check for NA}
is.na(survey_data)
colSums(is.na(survey_data))
```

`naniar` provides principled, tidy ways to summarise, visualise, and manipulate missing data with minimal deviations from the workflows in ggplot2 and tidy data. 

Resources: 

- https://r4ds.hadley.nz/missing-values 
- https://www.r-bloggers.com/2021/04/handling-missing-values-in-r/ 
- https://naniar.njtierney.com

```{r pct missing}
# Shows count AND percentage missing
naniar::miss_var_summary(survey_data) 
```

```{r missing plot}
# Visualize missing patterns
naniar::gg_miss_var(survey_data)
```

### What to do about them?

1.   Ignore
2.   Remove
3.   Replace

```{r}
# remove
data_dropna <- drop_na(data)
data_dropna
```

```{r}
# fill, but what does it do?
fill(data)
```

```{r}
which(data == 0, arr.ind = TRUE)
```







## Duplicates

```{r}
# check values in column, should match length if all unique
number_of_ages <- nrow(unique(data[, 'Age']))
number_of_ages
```

```{r}
nrow(data)
```

```{r}
# drop duplicates if exact
distinct(data)
```

## Data Types

```{r}
sapply(data, class)
```
