---
output: html_document
---

# Exploratory Data Analysis in R

Resource: https://r4ds.had.co.nz/tidy-data.html 

```{r setup, include=FALSE}
# Load libraries
library(ggplot2)
library(here)
library(skimr)
library(tidyverse)

# Set global theme for plots
theme_set(theme_minimal())
```

## Import data

`here` makes it easier to work with file paths by automatically detecting your project's root directory, so you don't have to worry about working directory issues or writing complex relative paths.

Resource: 

- https://here.r-lib.org
- https://www.r-bloggers.com/2024/05/easy-data-cleaning-with-the-janitor-package/

```{r import data}
# import the file using here and read_csv, expecting 89184 obs. of 84 variables
survey_data = readr::read_csv(here::here("data", "survey_results_public.csv")) %>% janitor::clean_names()
```

## View data

The following are some ways to examine the data!

```{r head}
# view the first few rows, default is 6
head(survey_data)
```

```{r tail}
# view the last few rows, default is 6
tail(survey_data)
```

```{r colnames}
# view column names
colnames(survey_data)
```

```{r number of rows}
# count number of rows
nrow(survey_data)
```

### Exercise 1

1.  Read your data into R and view it

## Select data

`dplyr` is part of the larger `tidyverse`, it provides convenient tools for the most common data manipulation tasks e.g., selecting rows/columns and filtering.

Resource: <https://cengel.github.io/R-data-wrangling/dplyr.html>

### Select col or row by position

```{r single row}
# select single row based on position
survey_data %>% 
  dplyr::slice(2) # R uses 1-based indexing
```

```{r single column}
# select single column based on position
survey_data %>% 
  dplyr::select(2)
```

```{r range of rows}
# select a range of rows based on position
survey_data %>% 
  dplyr::slice(2:5)
```

```{r specific rows}
# select specific rows based on position
survey_data %>% 
  dplyr::slice(c(2, 6, 11))
```

```{r range of columns}
# select a range of columns based on position
survey_data %>% 
  dplyr::select(2:5)
```

```{r specific cols}
# select specific columns based on position
survey_data %>% 
  dplyr::select(c(2, 6, 11))
```

```{r negative index}
# select last 10 rows
survey_data %>% 
  dplyr::slice_tail(n = 10)
```

### Select col by name

```{r}
# Select specific column by name 
survey_data %>% 
  dplyr:: select(RemoteWork)
```

## Filtering Data

Different ways to identify and count specific values/categories

### Specific conditions

```{r filter numeric}
# Only grab those with comp total greater than 500, 000
survey_data %>%
  filter(comp_total > 500000)
```

### Specific cateogries

```{r filter category}
# Only grab those from Australia
survey_data %>%
  filter(country == "Australia")
```

### Multiple conditions

```{r filter multiple}
# Only grab those from Australia with a comp total greater than 500, 0000
survey_data %>%
  filter(comp_total > 500000 & country == "Australia")
```

## Plotting Data

`ggplot2` is part of the `tidyverse` and it is a package that is dedicated to data visualization.

Resources:

-   <https://ggplot2.tidyverse.org/articles/ggplot2.html>
-   <https://r-graph-gallery.com/ggplot2-package.html>
-   <https://exts.ggplot2.tidyverse.org/gallery/>

```{r plot WorkExp}
# Start with basic histogram
ggplot(survey_data, aes(x = work_exp)) +
  geom_histogram(bins = 30, na.rm = TRUE)

# # Add colors and labels
# ggplot(survey_data, aes(x = WorkExp)) +
#   geom_histogram(bins = 30, 
#                 na.rm = TRUE,
#                 fill = "steelblue",
#                 color = "white") +
#   labs(title = "Distribution of Work Experience",
#        subtitle = "Stack Overflow Developer Survey 2023",
#        x = "Years of Work Experience",
#        y = "Count")
```

### Exercise 2

1.  Plot your own data
2.  Choose single column to plot

# Lunch Break 

## Reshaping data

Common problems: 
- Variable across multiple columns 
- Observations across multiple rows

Resource: https://cran.r-project.org/web/packages/tidyr/vignettes/pivot.html 

### Pivot longer

https://tidyr.tidyverse.org/reference/pivot_longer.html 

```{r pivot longer}
# making data longer by converting columns to rows
survey_data_long <- survey_data %>%
  pivot_longer(cols = -response_id, # keep ResponseId as identifier
               names_to = "variable",
               values_to = "value",
               values_transform = as.character) #need to correct format

# pivot_longer(survey_data, 
#              c('CompTotal', 'ConvertedCompYearly'), 
#              names_to = 'money', 
#              values_to = 'dollah')

dim(survey_data_long)
```

### Pivot wider

https://tidyr.tidyverse.org/reference/pivot_wider.html

```{r pivot wider}
survey_data_wide <- survey_data %>%
  pivot_wider(id_cols = response_id,
              names_from = age,
              values_from = -response_id)

dim(survey_data_wide)
```

### Exercise 3

1.  Convert your own data into a tidier format.


## Missing Data

Different types of missing data:

- NA values
- Empty strings
- Just whitespace
- Zeros (potential missing codes)
- Special codes e.g., "NA", "None", "null", "999"

```{r check for NA}
is.na(survey_data)
colSums(is.na(survey_data))
```

`naniar` provides principled, tidy ways to summarise, visualise, and manipulate missing data with minimal deviations from the workflows in ggplot2 and tidy data. 

Resources: 

- https://r4ds.hadley.nz/missing-values 
- https://www.r-bloggers.com/2021/04/handling-missing-values-in-r/ 
- https://naniar.njtierney.com

```{r pct missing}
# Shows count AND percentage missing
naniar::miss_var_summary(survey_data) 
```

```{r missing plot}
# Visualize missing patterns
naniar::gg_miss_var(survey_data)
```

### What to do about them?

1.   Ignore - Leave as is
2.   Remove - Completely remove or apply threshold
3.   Replace - Imputation 

#### Remove

```{r drop na}
# Remove rows with ANY missing values - not recommended because in most cases will remove everything
drop_na(survey_data)

# Remove with a certain threshold, in this case drop vars with >20% NAs
survey_data_dropped <- survey_data %>%
  dplyr::select(where(~mean(is.na(.)) < 0.7))

# Check which columns were dropped
setdiff(names(survey_data), names(survey_data_dropped))
```
#### Replace

On a case by cases basis can also:

- Impute mean/median for numerical
- Assign "Missing" or "Unknown" for categorical

```{r fill}
# Not too sure about fill here, ok for time series
survey_data_filled <- survey_data %>%
  fill(everything(), .direction = "down")

colSums(is.na(survey_data_filled))
```

Resource for summarise: https://dplyr.tidyverse.org/reference/summarise.html

```{r count zeros}
# Only show columns with zeros
survey_data %>%
  summarise(across(everything(), ~sum(. == 0, na.rm = TRUE))) %>%
  pivot_longer(everything(),
              names_to = "column",
              values_to = "zero_count") %>%
  filter(zero_count > 0) %>%  
  arrange(desc(zero_count))
```

## Duplicates

```{r get dupes}
# No dupes found
survey_data %>% janitor::get_dupes()

survey_data_unique <- survey_data %>% distinct()

setdiff(survey_data, survey_data_unique)
```

```{r check dupes examples}
# check the values in a column, if should be unique should match length of data
survey_data %>%
  summarise(unique_ages = n_distinct(age),
            total_rows = n())

survey_data %>%
  summarise(unique_ages = n_distinct(response_id),
            total_rows = n())
```

## Data Types

```{r glimpse or skim}
# a quick peek
glimpse(survey_data)

# with a little more detail
skimr::skim(survey_data)

#summary(survey_data)
#str(survey_data)
```

### Numeric

R doesn't distinguish between integers and floats, most numbers in R are "numeric" (double).

```{r}
num <- 3
class(num) #integer in python
typeof(num) #will show as dbl

float <- 3.14
class(float)  # Also "numeric"
typeof(float) # Also "double"

class(survey_data$response_id)
typeof(survey_data$response_id)

#check if numeric
is.numeric(survey_data$response_id)
is.double(survey_data$response_id)
is.integer(survey_data$response_id)

survey_data %>%
  select(where(is.numeric)) %>%
  summary()
```

### Categorical

```{r}
boolean <- FALSE 
class(boolean) #logical in R
typeof(boolean)

#there isn't any in our data
survey_data %>%
  dplyr::select(where(is.logical)) %>%
  summary()

#so let's make one, picking coding_activities
unique(survey_data$coding_activities)

# check how much contain "Hobby", still not logical
survey_data %>%
  mutate(hobby = str_detect(coding_activities, "Hobby")) %>%
  count(hobby, na.rm = TRUE) %>%
  mutate(pct = n/sum(n) * 100)

#remove NAs and only remove TRUE/FALSE
survey_data_hobby <- survey_data %>%
  filter(!is.na(coding_activities)) %>%  # Remove NAs first
  mutate(hobby = str_detect(coding_activities, "Hobby")) %>%
  dplyr::select(coding_activities, hobby)

class(survey_data_hobby$hobby) # now logical
```

```{r}
# what happens if we make something that isn't logical, logical?


```



### String

```{r}

```

### DateTime

```{r}

```

### Outliers

```{r}

```