---
output: html_document
---

# Exploratory Data Analysis in R

Resource: https://r4ds.had.co.nz/tidy-data.html 

```{r setup, include=FALSE}
# Load libraries
library(ggplot2)
library(here)
library(lubridate)
library(skimr)
library(tidyverse)
library(janitor)

# Set global theme for plots
theme_set(theme_minimal())
```

## Import data

`here` makes it easier to work with file paths by automatically detecting your project's root directory, so you don't have to worry about working directory issues or writing complex relative paths.

Resource: 

- https://here.r-lib.org
- https://www.r-bloggers.com/2024/05/easy-data-cleaning-with-the-janitor-package/

```{r import data}
# import the file using here and read_csv, expecting 89184 obs. of 84 variables
survey_data <- read_csv(here::here("data", "survey_results_public.csv")) %>% clean_names()
```

## View data

The following are some ways to examine the data!

```{r head}
# view the first few rows, default is 6
head(survey_data)
```

```{r tail}
# view the last few rows, default is 6
tail(survey_data)
```

```{r colnames}
# view column names
colnames(survey_data)
```

```{r number of rows}
# count number of rows
nrow(survey_data)
```

## Back to slides 
  
## Plotting Data

`ggplot2` is part of the `tidyverse` and it is a package that is dedicated to data visualization.

Resources:

-   <https://ggplot2.tidyverse.org/articles/ggplot2.html>
-   <https://r-graph-gallery.com/ggplot2-package.html>
-   <https://exts.ggplot2.tidyverse.org/gallery/>

```{r}
?ggplot
```


```{r plot WorkExp}
# Start with basic histogram
survey_data %>% 
    ggplot(data=survey_data, aes(x=work_exp)) + 
    geom_histogram()

survey_data %>%
    ggplot(data=survey_data, aes(x=age, y=work_exp)) + 
    geom_point()


ggplot(survey_data, aes(x = work_exp)) +
  geom_histogram(bins = 30, na.rm = TRUE)

# # Add colors and labels
# ggplot(survey_data, aes(x = WorkExp)) +
#   geom_histogram(bins = 30, 
#                 na.rm = TRUE,
#                 fill = "steelblue",
#                 color = "white") +
#   labs(title = "Distribution of Work Experience",
#        subtitle = "Stack Overflow Developer Survey 2023",
#        x = "Years of Work Experience",
#        y = "Count")
```

## Exercise 

Make a plot!
Use colnames to choose variables
What does this plot mean?

Put answers in google doc


## Select data

`dplyr` is part of the larger `tidyverse`, it provides convenient tools for the most common data manipulation tasks e.g., selecting rows/columns and filtering.

Resource: <https://cengel.github.io/R-data-wrangling/dplyr.html>

### Select col or row by position

```{r single row}
# select single row based on position
survey_data %>% 
  dplyr::slice(2) # R uses 1-based indexing
```

```{r single column}
# select single column based on position
survey_data %>% 
  dplyr::select(2)
```

```{r range of rows}
# select a range of rows based on position
survey_data %>% 
  dplyr::slice(2:5)
```

```{r specific rows}
# select specific rows based on position
survey_data %>% 
  dplyr::slice(c(2, 6, 11))
```

```{r range of columns}
# select a range of columns based on position
survey_data %>% 
  dplyr::select(2:5)
```

```{r specific cols}
# select specific columns based on position
survey_data %>% 
  dplyr::select(c(2, 6, 11))
```

```{r negative index}
# select last 10 rows
survey_data %>% 
  dplyr::slice_tail(n = 10)
```

### Select col by name

```{r}
# Select specific column by name 
survey_data %>% 
  dplyr:: select(RemoteWork)
```

```{r}
survey_data[10,]

survey_data %>%
  slice(10) %>%
  select(10)

```


## Exercise 
Select the data on your birthday! 
Months are columns
Days are rows

Put answer + code in google doc



## Filtering Data

Different ways to identify and count specific values/categories

### Specific conditions

```{r filter numeric}
# Only grab those with comp total greater than 500, 000
survey_data %>%
  filter(comp_total > 500000)
```

### Specific cateogries

```{r filter category}
# Only grab those from Australia
survey_data %>%
  filter(country == "Australia")
```

### Multiple conditions

```{r filter multiple}
# Only grab those from Australia with a comp total greater than 500, 0000
survey_data %>%
  filter(comp_total > 500000 & country == "Australia")
```

## Set data into the original df 
```{r}
survey_data$age <- 10

survey_data %>%
  mutate(age=ifelse(age>30, 'millenial', 'genz'))
```



# Lunch Break 

## Reshaping data

Common problems: 
- Variable across multiple columns 
- Observations across multiple rows

Resource: https://cran.r-project.org/web/packages/tidyr/vignettes/pivot.html 

### Pivot longer

https://tidyr.tidyverse.org/reference/pivot_longer.html 

```{r pivot longer}
# making data longer by converting columns to rows
survey_data_long <- survey_data %>%
  pivot_longer(cols = -response_id, # keep ResponseId as identifier
               names_to = "variable",
               values_to = "value",
               values_transform = as.character) #need to correct format

# pivot_longer(survey_data, 
#              c('CompTotal', 'ConvertedCompYearly'), 
#              names_to = 'money', 
#              values_to = 'dollah')

dim(survey_data_long)
```

### Pivot wider

https://tidyr.tidyverse.org/reference/pivot_wider.html

```{r pivot wider}
survey_data_wide <- survey_data %>%
  pivot_wider(id_cols = response_id,
              names_from = age,
              values_from = -response_id)

dim(survey_data_wide)
```

### Exercise 3

1.  Convert your own data into a tidier format.


## Missing Data

Different types of missing data:

- NA values
- Empty strings
- Just whitespace
- Zeros (potential missing codes)
- Special codes e.g., "NA", "None", "null", "999"

```{r check for NA}
is.na(survey_data)
colSums(is.na(survey_data))
```

`naniar` provides principled, tidy ways to summarise, visualise, and manipulate missing data with minimal deviations from the workflows in ggplot2 and tidy data. 

Resources: 

- https://r4ds.hadley.nz/missing-values 
- https://www.r-bloggers.com/2021/04/handling-missing-values-in-r/ 
- https://naniar.njtierney.com

```{r pct missing}
# Shows count AND percentage missing
naniar::miss_var_summary(survey_data) 
```

```{r missing plot}
# Visualize missing patterns
naniar::gg_miss_var(survey_data)
```

### What to do about them?

1.   Ignore - Leave as is
2.   Remove - Completely remove or apply threshold
3.   Replace - Imputation 

#### Remove

```{r drop na}
# Remove rows with ANY missing values - not recommended because in most cases will remove everything
drop_na(survey_data)

# Remove with a certain threshold, in this case drop vars with >20% NAs
survey_data_dropped <- survey_data %>%
  dplyr::select(where(~mean(is.na(.)) < 0.7))

# Check which columns were dropped
setdiff(names(survey_data), names(survey_data_dropped))
```
#### Replace

On a case by cases basis can also:

- Impute mean/median for numerical
- Assign "Missing" or "Unknown" for categorical

```{r fill}
# Not too sure about fill here, ok for time series
survey_data_filled <- survey_data %>%
  fill(everything(), .direction = "down")

colSums(is.na(survey_data_filled))
```

Resource for summarise: https://dplyr.tidyverse.org/reference/summarise.html

```{r count zeros}
# Only show columns with zeros
survey_data %>%
  summarise(across(everything(), ~sum(. == 0, na.rm = TRUE))) %>%
  pivot_longer(everything(),
              names_to = "column",
              values_to = "zero_count") %>%
  filter(zero_count > 0) %>%  
  arrange(desc(zero_count))
```

## Duplicates

```{r get dupes}
# No dupes found
survey_data %>% janitor::get_dupes()

survey_data_unique <- survey_data %>% distinct()

setdiff(survey_data, survey_data_unique)
```

```{r check dupes examples}
# check the values in a column, if should be unique should match length of data
survey_data %>%
  summarise(unique_ages = n_distinct(age),
            total_rows = n())

survey_data %>%
  summarise(unique_ages = n_distinct(response_id),
            total_rows = n())
```

## Data Types

Different ways to look at the overall data structure, classes and distributions

```{r classes}
# a quick peek
glimpse(survey_data)

# with a little more detail
skimr::skim(survey_data)

#summary(survey_data)
#str(survey_data)
```

### Numeric

R doesn't distinguish between integers and floats, most numbers in R are "numeric" (double).

```{r numeric}
num <- 3
class(num) #integer in python
typeof(num) #will show as dbl

float <- 3.14
class(float)  # Also "numeric"
typeof(float) # Also "double"

class(survey_data$response_id)
typeof(survey_data$response_id)

#check if numeric
is.numeric(survey_data$response_id)
is.double(survey_data$response_id)
is.integer(survey_data$response_id)

survey_data %>%
  select(where(is.numeric)) %>%
  summary()
```

### Logical

```{r logical}
boolean <- FALSE 
class(boolean) #logical in R
typeof(boolean)

#there isn't any in our data
survey_data %>%
  dplyr::select(where(is.logical)) %>%
  summary()

#so let's make one!
unique(survey_data$coding_activities)

# check how much contain "Hobby", still not logical
survey_data %>%
  mutate(hobby = str_detect(coding_activities, "Hobby")) %>%
  count(hobby, na.rm = TRUE) %>%
  mutate(pct = n/sum(n) * 100)

#remove NAs and only remove TRUE/FALSE
survey_data_hobby <- survey_data %>%
  filter(!is.na(coding_activities)) %>%  # Remove NAs first
  mutate(hobby = str_detect(coding_activities, "Hobby")) %>%
  dplyr::select(coding_activities, hobby)

class(survey_data_hobby$hobby) # now logical
```

What happens if we make something that isn't logical, logical?

```{r num to logical}
class(survey_data$converted_comp_yearly) #numeric

survey_data_comp_logical <- survey_data %>%
  select(converted_comp_yearly) %>%
  mutate(comp_logical = as.logical(converted_comp_yearly))

survey_data_comp_logical #anything not 0 counts as True
```

### Factors

Character - Stores text strings, sorts alphabetically, easier text manipulation
Factor - Stores cateogoriess as integers internally with predefined levels, better for modelling and for plotting with specific order

```{r factors}
class(survey_data$age) #character
unique(survey_data$age) 

# convert age to factor
survey_data_age <- survey_data %>%
  dplyr::select(age) %>%
  mutate(age_factor = factor(age)) %>%
  count(age_factor)

#check class and levels
class(survey_data_age$age_factor)
levels(survey_data_age$age_factor)

#order levels
survey_data_age <- survey_data %>%
  dplyr::select(age) %>%
  mutate(age_factor = factor(age, levels=c("Under 18 years old",
                                           "18-24 years old",
                                           "25-34 years old",
                                           "35-44 years old",
                                           "45-54 years old",
                                           "55-64 years old",
                                           "65 years or older",
                                           "Prefer not to say"))) %>%
  count(age_factor)

#check again
class(survey_data_age$age_factor)
levels(survey_data_age$age_factor)
```

What happens if we make something not a factor, a factor?

```{r num to factor}
class(survey_data$converted_comp_yearly) #numeric
unique(survey_data$converted_comp_yearly) 
levels(survey_data$converted_comp_yearly) #NULL because not factor

# Try to convert compensation to factor (probably not meaningful)
survey_data_comp_yearly <- survey_data %>%
  dplyr::select(converted_comp_yearly) %>%
  mutate(comp_factor = factor(converted_comp_yearly))

class(survey_data_comp_yearly$comp_factor)
levels(survey_data_comp_yearly$comp_factor) 
```

What is a string and how to check?

```{r string}
#take code_activities as an example
unique(survey_data$coding_activities)

# Check type of a single value
typeof(survey_data$coding_activities[1])  # "character"
class(survey_data$coding_activities[1])   # "character"

# Useful string operations with stringr
survey_data %>%
  filter(!is.na(coding_activities)) %>%
  transmute(
    original = coding_activities,
    length = str_length(coding_activities),
    uppercase = str_to_upper(coding_activities),
    n_activities = str_count(coding_activities, ";") + 1,
    first_activity = str_extract(coding_activities, "^[^;]+")) %>%
  view()

# # String manipulation examples
# survey_data %>%
#   filter(!is.na(coding_activities)) %>%
#   mutate(
#     # Split into list
#     activities_list = str_split(coding_activities, ";"),
#     # Count occurrences
#     n_semicolons = str_count(coding_activities, ";"),
#     # Extract patterns
#     first_word = str_extract(coding_activities, "\\w+"),
#     # Replace text
#     replaced = str_replace(coding_activities, "Hobby", "HOBBY"),
#     # Convert case
#     upper = str_to_upper(coding_activities),
#     lower = str_to_lower(coding_activities)
#   )
```

### DateTime

```{r ymd}
# Create a date from string
date_str <- "2020-01-01"
class(date_str) #character

# convert to dates in year-month-day using lubridate
date_time <- lubridate::ymd(date_str) 

class(date_time)  # date
typeof(date_time) #dbl

# Different ways to create dates
dates_example <- tibble(
  date_str = "2020-01-01",
  date_ymd = ymd(date_str),
  date_mdy = mdy("01-01-2020"),
  date_dmy = dmy("01-01-2020"),
  datetime = ymd_hms("2020-01-01 12:30:45"))

# Try to convert non-date data (like the compensation example)
# works but everything counted as milliseconds/seconds since 1970-01-01
survey_data %>%
  dplyr::select(converted_comp_yearly) %>%
  mutate(comp_as_date = as_datetime(converted_comp_yearly))
```

## Outliers

Identify outliers by plotting!

```{r histogram}
# look at the distribution
# Extremely high values obviously skewing
survey_data %>%
  filter(!is.na(comp_total)) %>% # 46.2% missing as NA so will get an error, lets remove
  ggplot(aes(x = comp_total)) +
  geom_histogram(bins=30) +
  labs(title = "Distribution of Compensation")

# Look at distribution with limit
survey_data %>%
  filter(!is.na(comp_total)) %>%
  ggplot(aes(x = comp_total)) +
  geom_histogram() +
  xlim(0, 1000000) +
  labs(title = "Distribution of Compensation (Limited Range)")

# Find high values
survey_data %>%
  filter(!is.na(comp_total)) %>%
  count(high_salary = comp_total > 1000000) %>%
  mutate(pct = n/sum(n) * 100)

# Create dataset without outliers
survey_data_filtered <- survey_data %>%
  filter(comp_total < 10000000) #removes 4521
```

```{r scatterplot}
# Scatter plot with work experience and outliers
ggplot(survey_data, aes(x = work_exp, y = comp_total)) +
  geom_point() +
  labs(title = "Compensation vs Work Experience (With Outliers)")

# Now without outliers
ggplot(survey_data_filtered, aes(x = work_exp, y = comp_total)) +
  geom_point() +
  labs(title = "Compensation vs Work Experience (Without Outliers)")

# Replace outliers with mean
survey_data_replaced <- survey_data %>%
  mutate(comp_replaced = case_when(comp_total > 10000000 ~ mean(comp_total, na.rm = TRUE),
                                   TRUE ~ comp_total))

# Replace outliers with mean and create plot
survey_data %>%
  mutate(comp_replaced = case_when(
    converted_comp_yearly > 10000000 ~ mean(converted_comp_yearly, na.rm = TRUE),
    TRUE ~ converted_comp_yearly
  )) %>%
  ggplot(aes(x = work_exp, y = comp_replaced)) +
  geom_point(alpha = 0.5) +  # alpha for transparency
  labs(
    title = "Compensation vs Work Experience",
    subtitle = "Outliers replaced with mean",
    x = "Years of Work Experience",
    y = "Yearly Compensation"
  ) +
  theme_minimal()
```

```{r}
# Compare statistics before and after replacement
summary(survey_data$converted_comp_yearly) # Original
summary(survey_data_replaced$comp_replaced) # After replacement

# Or if you want to compare both side by side
survey_data_replaced %>%
  select(converted_comp_yearly, comp_replaced) %>%
  summary()
```

### Basic export

```{r}
write_csv(survey_data_replaced, here("data", "survey_data_cleaned.csv"))
```

### Exercise 4

1. Check for outliers
2. Decide how to deal with them
3. Fix outliers
4. Save your data
