---
output: html_document
---

# Exploratory Data Analysis in R

## Import data

```{r} 
# import the file using here
data = here::here(readr::read_csv("data", "survey_results_public.csv"))
```

## View data

```{r}
# inspect the file
head(data)
```

```{r}
tail(data)
```

```{r}
# columns
colnames(data)
```

```{r}
# rows
row.names(data)
```

## Exercise
- Read your data into R and view it

## Select data
### Using position

```{r}
# single row
single_row <- data[3,]
single_row
```

```{r}
# single column
single_col <- data[,3]
single_col
```

```{r}
# range
many_rows <- data[3:6,]
many_rows
```

```{r}
many_cols <- data[,3:6]
many_cols
```

### Using labels

```{r}
remote_work_col <- data[, 'RemoteWork']
remote_work_col
```

### using condition

```{r}
remote_pos_response <- subset(data, RemoteWork == 'Remote')
remote_pos_response
```

## Plotting Data

```{r}
library(ggplot2)
ggplot(data=data, aes(x=WorkExp)) +
         geom_histogram()
```

## Exercise
- Plot your own data
- choose single column to plot

## Lunch Break

## Reformat data

Common problems
- variable across multiple columns 
- observation across multiple rows

```{r}
library(tidyverse)
pivot_longer(data, c('CompTotal', 'ConvertedCompYearly'), names_to = 'money', values_to = 'dollah')
```

```{r}
pivot_wider(data, names_from = Age, values_from = WorkExp)
```

## Exercise
- convert your own data into tidy format

## Clean data

### Missing Data
```{r}
# check for nulls
is.na(data)
```

```{r}
sum(is.na(data))
```

```{r}
summarise_all(data, ~ sum(is.na(.)))
```

```{r}
which(data == 999, arr.ind = TRUE)
```

```{r}
unique(single_col)
```

### What to do about them?
- Ignore
- remove 
- replace

```{r}
# remove
data_dropna <- drop_na(data)
data_dropna
```

```{r}
# fill, but what does it do?
fill(data)
```

```{r}
which(data == 0, arr.ind = TRUE)
```

## Duplicates

```{r}
# check values in column, should match length if all unique
number_of_ages <- nrow(unique(data[, 'Age']))
number_of_ages
```

```{r}
nrow(data)
```

```{r}
# drop duplicates if exact
distinct(data)
```

## Data Types

```{r}
sapply(data, class)
```
