---
output: html_document
---

# Exploratory Data Analysis in R

```{r setup, include=FALSE}
# Load libraries
library(here)
library(tidyverse)

# Set global theme for plots
theme_set(theme_minimal())
```

## Import data

`here` makes it easier to work with file paths by automatically detecting your project's root directory, so you don't have to worry about working directory issues or writing complex relative paths.

Resource: https://here.r-lib.org 

```{r import data} 
# import the file using here and read_csv, expecting 89184 obs. of 84 variables
survey_data = readr::read_csv(here::here("data", "survey_results_public.csv"))
```

## View data

The following are some ways to examine the data!

```{r head}
# view the first few rows, default is 6
head(survey_data)
```

```{r tail}
# view the last few rows, default is 6
tail(survey_data)
```

```{r colnames}
# view column names
colnames(survey_data)
```

```{r number of rows}
# count number of rows
nrow(survey_data)
```

## Exercise
- Read your data into R and view it

## Select data

`dplyr` is part of the larger `tidyverse`, it provides convenient tools for the most common data manipulation tasks e.g., selecting rows/columns and filtering.

Resource: https://cengel.github.io/R-data-wrangling/dplyr.html

### Select column or row by position

```{r}
# select single row based on position
survey_data %>% 
  dplyr::slice(2) # R uses 1-based indexing
```

```{r}
# select single column based on position
survey_data %>% 
  dplyr::select(2)
```

```{r}
# select a range of rows based on position
survey_data %>% 
  dplyr::slice(2:5)
```

```{r}
# select specific rows based on position
survey_data %>% 
  dplyr::slice(c(2, 6, 11))
```

```{r} 
# select a range of columns based on position
survey_data %>% 
  dplyr::select(2:5)
```

```{r}
# select specific columns based on position
survey_data %>% 
  dplyr::select(c(2, 6, 11))
```

```{r}
# Negative indexing, select last 10 rows
survey_data %>% 
  dplyr::slice_tail(n = 10)
```

### Select column by name

```{r}
# Select specific column by name 
survey_data %>% 
  dplyr:: select(RemoteWork)
```

## Filtering Data

Different ways to identify and count specific values/categories

### Specific conditions

```{r}
# Only grab those with comp total greater than 500, 000
survey_data %>%
  filter(CompTotal > 500000)
```

### Specific cateogries

```{r}
# Only grab those from Australia
survey_data %>%
  filter(Country == "Australia")
```

### Multiple conditions

```{r}
# Only grab those from Australia with a comp total greater than 500, 0000
survey_data %>%
  filter(CompTotal > 500000 & Country == "Australia")
```





## Plotting Data

```{r}
library(ggplot2)
ggplot(data=data, aes(x=WorkExp)) +
         geom_histogram()
```

## Exercise
- Plot your own data
- choose single column to plot

## Lunch Break

## Reformat data

Common problems
- variable across multiple columns 
- observation across multiple rows

```{r}
library(tidyverse)
pivot_longer(data, c('CompTotal', 'ConvertedCompYearly'), names_to = 'money', values_to = 'dollah')
```

```{r}
pivot_wider(data, names_from = Age, values_from = WorkExp)
```

## Exercise
- convert your own data into tidy format

## Clean data

### Missing Data
```{r}
# check for nulls
is.na(data)
```

```{r}
sum(is.na(data))
```

```{r}
summarise_all(data, ~ sum(is.na(.)))
```

```{r}
which(data == 999, arr.ind = TRUE)
```

```{r}
unique(single_col)
```

### What to do about them?
- Ignore
- remove 
- replace

```{r}
# remove
data_dropna <- drop_na(data)
data_dropna
```

```{r}
# fill, but what does it do?
fill(data)
```

```{r}
which(data == 0, arr.ind = TRUE)
```

## Duplicates

```{r}
# check values in column, should match length if all unique
number_of_ages <- nrow(unique(data[, 'Age']))
number_of_ages
```

```{r}
nrow(data)
```

```{r}
# drop duplicates if exact
distinct(data)
```

## Data Types

```{r}
sapply(data, class)
```
